---
title: "Reproducible Research: Peer Assessment 1"
author: "Joe Zendle"
date: "Jan, 5 2016"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

library(dplyr, quietly = TRUE)
unzip(zipfile = "activity.zip",overwrite = TRUE)
csv <- read.csv("activity.csv")

# add date object column
csv <- mutate(csv, dateDate = as.Date(date))

str(csv)
summary(csv)

# how many NA's?
sum(is.na(csv$steps))
# many

# how many zeros?
sum(csv$steps[csv$steps == 0], na.rm=T)

# none

```
## What is mean total number of steps taken per day?

From problem statement:

1) Calculate the total number of steps taken per day.

```{r}
byDate <- group_by(csv, dateDate)

stepsByDate <- summarise(byDate, numSteps = sum(steps, na.rm = TRUE))
```

2) Make a histogram of the total number of steps taken each day.

```{r}
library(ggplot2)
qplot(numSteps, data=stepsByDate, binw=1500)

```

3) Calculate and report the mean and median of the total number of steps taken per day.

```{r}

# since there are no zeros in the original data, not fare to count NA's as zero
# so calculate mean with and without zeros (zeros come from sum() operation above)

mean( filter(stepsByDate, numSteps > 0)$numSteps)
mean( stepsByDate$numSteps)

# and corresponding standard deviations
sd(filter(stepsByDate, numSteps > 0)$numSteps) 
sd(stepsByDate$numSteps) 

```

## What is the average daily activity pattern?

From problem statement:

1) Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
byInterval <- group_by(csv, interval)

stepsByInterval <- summarise(byInterval, avgSteps = mean(steps, na.rm = TRUE))

# interval number are in  HHmm format so make them nicer on x-axis
v1 <- c(0,600,1200,1800,2300)
v2 <- c("12am","6am","12pm", "6pm","11pm")
# dev.off()
plot(x=stepsByInterval$interval,
     y=stepsByInterval$avgSteps,
     xlab="5 minute interval",
     ylab="num steps",
     main="Number of Steps by Daily 5 Minute Interval",
     lwd=3,
     col="red",
     type="l",
     xaxt="n") # build the x-axis ticks in next step
axis(side = 1, 
     at = v1, 
     labels = v2,
     tck=-.05)

```

2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

# find the value of interval corresponding with max value of average steps

filter( stepsByInterval, stepsByInterval$avgSteps == max(stepsByInterval$avgSteps))$int
```

From this calculation, the interval with the largest number of steps is between 8:35-8:40am.

## Imputing missing values

From problem statement:

1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
# how many NA's?
sum(is.na(csv$steps))
```

2) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I will use the mean of all intervals for the value of the NA's in the data:
```{r}
mnIntervals <- mean(csv$steps,na.rm=T)
mnIntervals
```

3) Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
```


4) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
```
From problem statement: 

1) Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
```

2) Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
```