---
title: "Reproducible Research: Peer Assessment 1"
author: "Joe Zendle"
date: "Jan, 5 2016"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

library(dplyr, quietly = TRUE)
unzip(zipfile = "activity.zip",overwrite = TRUE)
csv <- read.csv("activity.csv")

# add date object column
csv <- mutate(csv, dateDate = as.Date(date))

str(csv)
summary(csv)

# how many NA's?
sum(is.na(csv$steps))
# many

# how many zeros?
sum(csv$steps[csv$steps == 0], na.rm=T)

# none

```
## What is mean total number of steps taken per day?

From problem statement:

1) Calculate the total number of steps taken per day.

```{r}
byDate <- group_by(csv, dateDate)

stepsByDate <- summarise(byDate, numSteps = sum(steps, na.rm = TRUE))
```

2) Make a histogram of the total number of steps taken each day.

```{r}
library(ggplot2)
qplot(numSteps, data=stepsByDate, binw=1500)

```

3) Calculate and report the mean and median of the total number of steps taken per day.

```{r}

# since there are no zeros in the original data, not fare to count NA's as zero
# so calculate mean with and without zeros (zeros come from sum() operation above)

mean( filter(stepsByDate, numSteps > 0)$numSteps)
mean( stepsByDate$numSteps)

# and corresponding standard deviations
sd(filter(stepsByDate, numSteps > 0)$numSteps) 
sd(stepsByDate$numSteps) 

```

## What is the average daily activity pattern?

From problem statement:

1) Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
